---
title: "Uber_Data_Analysis_Project"
output: html_document
date: "2024-02-28"
---



#### Loading the necessary libraries.
```{r , warning=FALSE}

library(tidyverse)

library(lubridate)


```






#### Reading the data of 6 months starting from April to September into its respective dataframes.
```{r}

april_data <- read_csv("/Users/sharadraina/Desktop/archive/Project_02/Uber-dataset/uber-raw-data-apr14.csv" , show_col_types = FALSE)

may_data <- read_csv("/Users/sharadraina/Desktop/archive/Project_02/Uber-dataset/uber-raw-data-may14.csv" , show_col_types = FALSE)

june_data <- read_csv("/Users/sharadraina/Desktop/archive/Project_02/Uber-dataset/uber-raw-data-jun14.csv" , show_col_types = FALSE)

july_data <- read_csv("/Users/sharadraina/Desktop/archive/Project_02/Uber-dataset/uber-raw-data-jul14.csv" , show_col_types = FALSE)

august_data <- read_csv("/Users/sharadraina/Desktop/archive/Project_02/Uber-dataset/uber-raw-data-aug14.csv" , show_col_types = FALSE)

september_data <- read_csv("/Users/sharadraina/Desktop/archive/Project_02/Uber-dataset/uber-raw-data-sep14.csv" , show_col_types = FALSE)

head(september_data , 3)

```


##### All the dataframes convey the similar information w.r.t Date/Time and Latitude and Longitude and Base. The date/time column is present in character format and needs to be tidyed in appropriate datatype.




#### I am using rbind function to append the data as all the dataframes have same number of columns and represent a time frame from April-September which will make it easier to analyse when joined row wise.
```{r}
# Joining all the dataframes into one named data_2014.

data_2014 <- rbind(april_data , may_data,june_data,july_data,august_data,september_data)




cat("The final dataframe consists of" , dim(data_2014) [1] , "rows","and" , dim(data_2014)[2] , "columns")


```



#### Tidying the date/time column.
```{r}

# Using as.PosixCT function to convert datatype of date/time column and formatting it in appropriate ymd format.
data_2014$`Date/Time` <- as.POSIXct(data_2014$`Date/Time` , format = "%m/%d/%Y %H:%M:%S")

head(data_2014$`Date/Time` , 5)
```



```{r}

#Creating a separate column which records Time in the form of hours:minutes:seconds
data_2014$Time <- format(as.POSIXct(data_2014$`Date/Time`, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")

head(data_2014$Time , 3 )


```



#### Creating variables that store values from Hour, Minute and Second which can then be used to visualize data for particular time intervals. I will use Lubridate package for creating/handling date/time variables.
```{r}

data_2014$Hour <- hour(data_2014$`Date/Time`)

data_2014$Minutes <- minute(data_2014$`Date/Time`)

data_2014$Seconds <- second(data_2014$`Date/Time`)

head(data_2014 , 3)


```



### Creating values for month, day , weekday and year which can be used to visualize data for particular days.
```{r}

data_2014$Day <- day(data_2014$`Date/Time`)

data_2014$Weekday <- wday(data_2014$`Date/Time`)

data_2014$Month <- month(data_2014$`Date/Time`)

data_2014$Year <- year(data_2014$`Date/Time`)

data_2014$Monthname <- month(data_2014$`Date/Time` , label = TRUE)

data_2014$Dayname  <- wday(data_2014$`Date/Time` , label = TRUE)

head(data_2014)

```



```{r}

#Dropping NA values
data_2014 =  drop_na(data_2014)


```





#### Plotting trips by the hour
```{r}

data_2014 %>%
  
  group_by(Hour) %>%
  
  summarise(Total = n()) %>%
  
  ggplot(aes(factor(Hour) ,Total / 1000 )) +
  
  scale_y_continuous(labels = function(x) paste0(x , "K")) +
  
   geom_bar(stat = "identity" , fill = "red") +
  
  theme_classic() +
  
  labs(title = "Total trips by each Hour (24 Hour)" , 
       
       x = "Hour of the Day" , 
       
       y = "Total trips")

  
  
 



```


##### The hours between 2:00 am and 4:00 am experience the lowest number of trips, indicating reduced demand during these late-night hours. Companies can adjust their driver scheduling and resource allocation during this time to optimize efficiency and minimize costs.

##### A steady increase in trips is observed from 6:00 am to 8:00 am, likely due to commuters traveling to work or other daytime activities. This morning rush hour signifies a peak in demand, suggesting the need for adequate driver availability to meet the increased transportation needs of passengers during this time. 

##### Max trips throughtout the day occur in the evening around 7pm so hours feom 5-9 can be considered the busiest for these drivers. Hence maximum demand that gets generated is during evening hours.

##### Overall, 2 peaks are witnessed throughout the day: One in morning during office hours and the second during evening time when usually office hours for the day would finish.

##### With a clear understanding of peak hours and low-demand periods, businesses can prioritize service availability during peak times to ensure timely pickups and shorter wait times for customers. This leads to improved customer satisfaction and loyalty, ultimately driving repeat business and positive word-of-mouth referrals.

##### Armed with knowledge of peak demand periods, businesses can design targeted incentive programs to encourage driver availability during these times. This can include bonuses, promotions, or rewards for drivers who operate during high-demand hours, thereby ensuring a sufficient supply of drivers to meet customer demand.





####Plotting trips by Month
```{r}

data_2014 %>%
  
  group_by(Monthname ) %>%
  
  summarise(Total = n()) %>%
  
  ggplot(aes( Monthname, Total / 1000 )) +
  
  scale_y_continuous(labels = function(x)  paste0(x , "K")) +
  
  geom_bar(stat = "identity" , fill = "red") +
  
  theme_classic() +
  
  labs(title = "Trips for each month (Apr - Sep)" , 
       
       x = "Month" , 
       
       y = "Total Trips") 



```

##### The peak months of August and September indicate periods of high demand, presenting opportunities for increased revenue generation. Businesses can capitalize on these peak periods by implementing surge pricing or targeted promotional campaigns to maximize profits.

##### Understanding the seasonal variation in demand allows businesses to allocate resources more effectively. For example, during peak months, they can ensure sufficient availability of drivers and vehicles to meet customer demand, while during off-peak months, they can optimize resource utilization to minimize costs.

##### Analyzing the reasons behind the fluctuation in demand can provide insights into customer behavior. Businesses can use this information to tailor their services and offerings to better meet customer needs and preferences, ultimately enhancing customer satisfaction and loyalty.

#####  If there is a significant increase in demand in a particular region during a specific month, businesses can consider expanding their operations to capitalize on this opportunity.

##### Each month's performance provides valuable insights into the demand for Uber services, which can help in strategic planning, resource allocation, and marketing efforts.



```{r}

data_2014 %>%
  
  group_by(Monthname , Hour) %>%
  
  summarise(Total = n()) %>%
  
  ggplot(aes( Hour,Total / 1000 , fill =  Monthname)) +
  
  scale_y_continuous(labels = function(x) paste0(x , "K")) +
  
    theme_classic() +
  
  labs(title = "Trips for each Hour by month (Apr - Sep)" , 
       
       x = "Hour of the Day" , 
       
       y = "Total Trips") +


  
  geom_bar(stat = "identity")


```

##### Combining the above 2 visualizations to see which months and which hours of the day contribute to maximum uber trips.






```{r}

data_2014 %>%
  
  group_by(Monthname , Dayname) %>%
  
  summarise(Total = n()) %>%
  
  ggplot(aes( Monthname, Total / 1000 , fill = Dayname  )) +
  
    scale_y_continuous(labels = function(x) paste0(x , "K")) +
  
  geom_bar(stat = "identity" , position = "dodge") +
  
      theme_classic() +
  
  labs(title = "Trips for month (Apr - Sep) by Day of the Week" , 
       
       x = "Month" , 
       
       y = "Total Trips") 











```

##### The highest number of Uber trips occurring in August and September suggests that these months experience the highest demand for ride-sharing services. This could be attributed to factors such as summer vacations, events, or increased outdoor activities during these months.

##### The observation that Thursday, Friday, and Saturday consistently have the highest number of trips for each month implies that these days are peak days for ride-sharing services. This pattern aligns with typical trends in consumer behavior, where people may require more transportation options for social events, outings, or weekend leisure activities.

##### The insights from this visualization present opportunities for Uber and other ride-sharing companies to strategically allocate resources and tailor promotional activities to capitalize on peak demand periods, particularly during the summer months and on weekends. This could involve targeted marketing campaigns, special promotions, or incentives aimed at increasing rider engagement and driver availability during these high-demand periods.


































